---
title: "FACTOR ANALYSIS IN R"
author: "Marjorie Blanco"
date: "12/20/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(psych)
```

## 

```{r}
gcbs <- readRDS("~/GitHub/datacamp/factor/GCBS_data.rds")
str(gcbs)
EFA_model <- fa(gcbs)
fa.diagram(EFA_model)
EFA_model$loadings
```

Inspecting your dataset

```{r}
describe(gcbs)
```

Splitting the dataset

```{r}
N <- nrow(gcbs)
indices <- seq(1, N)
indices_EFA <- sample(indices, floor((.5*N))) 
indices_CFA <- indices[!(indices %in% indices_EFA)]
gcbs_EFA <- gcbs[indices_EFA, ] 
gcbs_CFA <- gcbs[indices_CFA, ]
```

Inspecting the halves

```{r}
group_var <- vector("numeric", nrow(gcbs)) 
group_var[indices_EFA] <- 1 
group_var[indices_CFA] <- 2
```

Inspecting the halves

```{r}
gcbs_grouped <- cbind(gcbs, group_var)
describeBy(gcbs_grouped, group = group_var) 
statsBy(gcbs_grouped, group = "group_var")
```

Correlations

```{r}
lowerCor(gcbs)
```

Testing Correlations' Significance: P-values

```{r}
corr.test(gcbs, use = "pairwise.complete.obs")$p
```

Testing Correlations' Significance: Confidence Intervals

```{r}
corr.test(gcbs, use = "pairwise.complete.obs")$ci
```

Coefficient Alpha

```{r}
alpha(gcbs)
```

Split-Half reliability

```{r}
splitHalf(gcbs)
```

```{r}
head(bfi)
names(bfi)
```

Setup: Split your dataset

```{r}
# Establish two sets of indices to split the dataset N <- nrow(bfi)
indices <- seq(1, N)
indices_EFA <- sample(indices, floor((.5*N)))
indices_CFA <- indices[!(indices %in% indices_EFA)]
# Use those indices to split the dataset into halves for your EFA and CFA 
bfi_EFA <- bfi[indices_EFA, ]
bfi_CFA <- bfi[indices_CFA, ]
head(bfi_EFA)
head(bfi_CFA)
```

Calculate the correlation matrix

```{r}
# Calculate the correlation matrix first
bfi_EFA_cor <- cor(bfi_EFA, use = "pairwise.complete.obs")
```

Eigenvalues

```{r}
# Then use that correlation matrix to calculate eigenvalues 
eigenvals <- eigen(bfi_EFA_cor)
# Look at the eigenvalues returned 
eigenvals$values
```

Scree Plots

```{r}
# Calculate the correlation matrix first
bfi_EFA_cor <- cor(bfi_EFA, use = "pairwise.complete.obs")
# Then use that correlation matrix to create the scree plot 
scree(bfi_EFA_cor, factors = FALSE)
```

Running a multidimensional EFA

```{r}
# Run the EFA with six factors (as indicated by your scree plot) 
EFA_model <- fa(bfi_EFA, nfactors = 6)
# View results from the model object 
EFA_model
```

Factor loadings

```{r}
EFA_model$loadings
```

Factor scores

```{r}
head(EFA_model$scores)
```

Finding the fit statistics


```{r}
# Run the EFA with six factors (as indicated by your scree plot) 
EFA_model <- fa(bfi_EFA, nfactors = 6)
# View results from the model object 
EFA_model
```

Relative model fit

```{r}
# Run each theorized EFA on your dataset
bfi_theory <- fa(bfi_EFA, nfactors = 5) 
bfi_eigen <- fa(bfi_EFA, nfactors = 6)
# Compare the BIC values 
bfi_theory$BIC 
bfi_eigen$BIC
bfi_theory$BIC
bfi_eigen$BIC
```

Using the wrapper function to set up a CFA

```{r}
EFA_syn <- structure.sem(EFA_model)
EFA_syn
```

Creating CFA syntax from your theory

```{r}
# Set up syntax specifying which items load onto each factor
theory_syn_eq <- "
AGE: A1, A2, A3, A4, A5 
CON: C1, C2, C3, C4, C5 
EXT: E1, E2, E3, E4, E5 
NEU: N1, N2, N3, N4, N5 
OPE: O1, O2, O3, O4, O5 "
theory_syn <- cfa(text = theory_syn_eq, reference.indicators = FALSE)
```
Relationships between variables and factors

```{r}
theory_syn
```

Running the CFA

```{r}
#Use the sem() function to run a CFA 
theory_CFA <- sem(theory_syn, data = bfi_CFA)
```

Interpreting the output

```{r}
summary(theory_CFA)
```

Changing the Options

```{r}
options(fit.indices = c("CFI", "GFI", "RMSEA", "BIC"))
```

Absolute Model Fit

```{r}
summary(theory_CFA)
```

Relative Fit

```{r}
summary(theory_CFA)
summary(theory_CFA)$BIC
```

Relative Fit: Comparing Models

```{r}
summary(theory_CFA)$BIC
# Run a CFA using the EFA syntax you created earlier 
EFA_CFA <- sem(EFA_syn, data = bfi_CFA)
summary(EFA_CFA)$BIC
```

Differences in factor loadings

```{r}
# View the first five rows 
EFA_model$loadings[1:5,]
```

Differences in factor loadings

```{r}
# View the first five loadings from the CFA estimated from the EFA results 
summary(EFA_CFA)$coeff[1:5,]
```

Comparing factor loadings

```{r}
# View the first five loadings from the CFA estimated from the EFA results 
summary(EFA_CFA)$coeff[1:5,]

# View the first five rows of the EFA loadings 
EFA_model$loadings[1:5,]
```

Differences in factor scores

```{r}
# Extracting factor scores from the EFA model 
EFA_scores <- EFA_model$scores
# Calculate factor scores for the EFA dataset 
CFA_scores <- fscores(EFA_CFA, data = bfi_EFA)
```

Differences in factor scores, visualized

```{r}
plot(density(EFA_scores[,1], na.rm = TRUE), xlim = c(-3, 3), ylim = c(0, 1), col = "blue")
lines(density(CFA_scores[,1], na.rm = TRUE), xlim = c(-3, 3), ylim = c(0, 1), col = "red")
```

Adding loadings to the syntax

```{r}
summary(theory_CFA)
```

