---
title: "Fundamentals of Bayesian Data Analysis in R"
subtitle: "How does Bayesian inference work?"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(purrr)
library(tibble)
library(forcats)
library(ggplot2)
library(ggjoy)
```

```{r}
prop_model <- function (data = c(), prior_prop = c(1, 1), n_draws = 10000, 
    show_plot = TRUE) 
{
    data <- as.logical(data)
    proportion_success <- c(0, seq(0, 1, length.out = 100), 1)
    data_indices <- round(seq(0, length(data), length.out = min(length(data) + 
        1, 20)))
    post_curves <- map_dfr(data_indices, function(i) {
        value <- ifelse(i == 0, "Prior", ifelse(data[i], "Success", 
            "Failure"))
        label <- paste0("n=", i)
        probability <- dbeta(proportion_success, prior_prop[1] + 
            sum(data[seq_len(i)]), prior_prop[2] + sum(!data[seq_len(i)]))
        probability <- probability/max(probability)
        data_frame(value, label, proportion_success, probability)
    })
    post_curves$label <- fct_rev(factor(post_curves$label, levels = paste0("n=", 
        data_indices)))
    post_curves$value <- factor(post_curves$value, levels = c("Prior", 
        "Success", "Failure"))
    p <- ggplot(post_curves, aes(x = proportion_success, y = label, 
        height = probability, fill = value)) + geom_joy(stat = "identity", 
        color = "white", alpha = 0.8, panel_scaling = TRUE, size = 1) + 
        scale_y_discrete("", expand = c(0.01, 0)) + scale_x_continuous("Underlying proportion of success") + 
        scale_fill_manual(values = hcl(120 * 2:0 + 15, 100, 65), 
            name = "", drop = FALSE, labels = c("Prior   ", "Success   ", 
                "Failure   ")) + theme_light(base_size = 18) + 
        theme(legend.position = "top")
    if (show_plot) {
        print(p)
    }
    invisible(rbeta(n_draws, prior_prop[1] + sum(data), prior_prop[2] + 
        sum(!data)))
}
```

```{r}
# The generative zombie drug model

# Parameters
prop_success <- 0.15
n_zombies <- 13

# Simulating data
data <- c()
for(zombie in 1:n_zombies) {
  data[zombie] <- runif(1, min = 0, max = 1) < prop_success
}
data <- as.numeric(data)
data

prop_model(data)
```


```{r}
# The generative zombie drug model

# Parameters
prop_success <- 0.42
n_zombies <- 100

# Simulating data
data <- c()
for(zombie in 1:n_zombies) {
  data[zombie] <- runif(1, min = 0, max = 1) < prop_success
}
data <- as.numeric(data)
data

data <- sum(data)
data
```

```{r}
# Try out rbinom
rbinom(n = 1, size = 100, prob = .42)
# Try out rbinom
rbinom(n = 200, size = 100, prob = .42)
```

```{r}
cured_zombies <- rbinom(n = 100000, size = 100, prob = 0.07)
hist(cured_zombies)
```
How many visitors could your site get?

To get more visitors to your website you are considering paying for an ad to be shown 100 times on a popular social media site. According to the social media site, their ads get clicked on 10% of the time.

```{r}
# Fill in the parameters
n_samples <- 100000
n_ads_shown <- 100
proportion_clicks <- 0.1
n_visitors <- rbinom(n_samples, size = n_ads_shown, 
                     prob = proportion_clicks)

# Visualize n_visitors
hist(n_visitors)
```
Question
You would like the ad campaign to result in at least 5 visitors to your site.

Eyeballing the plot you just produced, what is the probability you will get 5 or more visitors because of the ad?

90%

```{r}
n_samples <- 100000
n_ads_shown <- 100
proportion_clicks <- 0.1
n_visitors <- rbinom(n_samples, size = n_ads_shown, 
                     prob = proportion_clicks)

# Visualize n_visitors
hist(n_visitors)


proportion_clicks <- runif(n = 6, min = 0.0, max = 1.0)
n_clicks <- rbinom(n = 6, size = 100, proportion_clicks)
# Visualize n_visitors
hist(n_visitors)
```
Adding a prior to the model

```{r}
# Update proportion_clicks
n_samples <- 100000
n_ads_shown <- 100
proportion_clicks <- runif(n = n_samples, min = 0.0, max = 0.2)
n_visitors <- rbinom(n = n_samples, size = n_ads_shown, prob = proportion_clicks)

# Visualize the results
hist(proportion_clicks)
hist(n_visitors)
```
Question

This looks very different from the histogram of n_visitors we got in the last exercise when proportion_clicks was exactly 0.1. With the added uncertainty in proportion_clicks the uncertainty over the number of visitors we 'll get also increased.

Eyeballing the plot you just produced, what is the probability you will get 5 or more visitors because of the ad?

70%

```{r}
prior <- data.frame(proportion_clicks, n_visitors)
prior
```

```{r}
# Create the prior data frame
prior <- data.frame(proportion_clicks, n_visitors)


# Create the posterior data frame
head(prior)

posterior <- prior[prior$n_visitors == 13, ]
hist(posterior$proportion_clicks)
```


Question
This doesn't look at all like the uniform distribution between 0.0 and 0.2 we put into proportion_clicks before. The whole distribution of samples now represent the posterior (after the data) probability distribution over what proportion_clicks could be.

Looking at the probability distribution over proportion_clicks what does the model know about the underlying proportion of visitors clicking on the ad?

It's likely between 7% and 19%.

```{r}
# Assign posterior to a new variable called prior
prior <-  posterior
# Take a look at the first rows in prior
head(prior, n =10)
# Replace prior$n_visitors with a new sample and visualize the result
n_samples <-  nrow(prior)
n_ads_shown <- 100
prior$n_visitors <- rbinom(n_samples, size = n_ads_shown,
                           prob = prior$proportion_clicks)
hist(prior$n_visitors)

# Calculate the probability that you will get 5 or more visitors
sum(prior$n_visitors >= 5)/length(prior$n_visitors)
```

Question
According to the new model, what is the probability of getting five or more visitors?

99%


