---
title: "Fundamentals of Bayesian Data Analysis in R"
subtitle: "Why use Bayesian Data Analysis?"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Explore using the rbeta function
beta_sample <- rbeta(n = 1000000, shape1 = 1, shape2 = 1)

# Visualize the results
hist(beta_sample)

# Explore using the rbeta function
beta_sample <- rbeta(n = 1000000, shape1 = 100, shape2 = 100)

# Visualize the results
hist(beta_sample)

# Explore using the rbeta function
beta_sample <- rbeta(n = 1000000, shape1 = 100, shape2 = 20)

# Visualize the results
hist(beta_sample)
```

Pick the prior that best captures the information
The new information you got from the social media company was:

Most ads get clicked on 5% of the time, but for some ads it is as low as 2% and for others as high as 8%.

There are many different probability distributions that one can argue captures this information.

Out of the four Beta distribution shown below, which captures this information the best?

```{r}
# Explore using the rbeta function
beta_sample <- rbeta(n = 1000000, shape1 = 3, shape2 = 12)

# Visualize the results
hist(beta_sample)

# Explore using the rbeta function
beta_sample <- rbeta(n = 1000000, shape1 = 15, shape2 = 75)

# Visualize the results
hist(beta_sample)


# Explore using the rbeta function
beta_sample <- rbeta(n = 1000000, shape1 = 5, shape2 = 95)

# Visualize the results
hist(beta_sample)

# Explore using the rbeta function
beta_sample <- rbeta(n = 1000000, shape1 = 30, shape2 = 10)

# Visualize the results
hist(beta_sample)
```

c. 
Look for a distribution that roughly covers the range 2% to 8%, that is, 0.02 to 0.08.

```{r}
n_draws <- 100000
n_ads_shown <- 100

# Change the prior on proportion_clicks
proportion_clicks <- 
  rbeta(n_draws, shape1 = 5, shape2 = 95)
n_visitors <- 
  rbinom(n_draws, size = n_ads_shown, 
         prob = proportion_clicks)
prior <- 
  data.frame(proportion_clicks, n_visitors)
posterior <- 
  prior[prior$n_visitors == 13, ]

# This plots the prior and the posterior in the same plot
par(mfcol = c(2, 1))
hist(prior$proportion_clicks, 
     xlim = c(0, 0.25))
hist(posterior$proportion_clicks, 
     xlim = c(0, 0.25))
```
Fit the model using another dataset
Let's fit the binomial model to both the video ad data (13 out of 100 clicked) and the new text ad data (6 out of a 100 clicked).

```{r}
n_draws <- 100000
n_ads_shown <- 100
proportion_clicks <- runif(n_draws, min = 0.0, max = 0.2)
n_visitors <- rbinom(n = n_draws, size = n_ads_shown, 
                     prob = proportion_clicks)
prior <- data.frame(proportion_clicks, n_visitors)

# Create the posteriors for video and text ads
posterior_video <- prior[prior$n_visitors == 13, ]
posterior_text <- prior[prior$n_visitors == 6, ]

# Visualize the posteriors
hist(posterior_video$proportion_clicks, xlim = c(0, 0.25))
hist(posterior_text$proportion_clicks, xlim = c(0, 0.25))
```

Question
Looking at the histogram of posterior_text what can be said about the value of the proportion of clicks for the text ad?
It's likely between 0.03 and 0.13.

```{r}
posterior <- data.frame(
    video_prop = posterior_video$proportion_clicks[1:4000],
    text_prop  = posterior_text$proportion_click[1:4000])
    
# Calculate the posterior difference: video_prop - text_prop
posterior$prop_diff <- posterior$video_prop - posterior$text_prop

# Visualize prop_diff
 hist(posterior$prop_diff, xlim = c(0, 0.25))

median(posterior$prop_diff)
# Summarize prop_diff
sum(posterior$prop_diff > 0) / length(posterior$prop_diff)
```

Question
Given the model and the data, what is the probability that the video ad is better than the text ad? (Here better means having a higher proportion of clicks.)


95%

```{r}
video_cost <- 0.25
text_cost <- 0.05
visitor_spend <- 2.53

posterior$video_profit <- 
  posterior$video_prop * visitor_spend - video_cost
posterior$text_profit <-
  posterior$text_prop * visitor_spend - text_cost
posterior$profit_diff <- posterior$video_profit - posterior$text_profit

hist(posterior$video_profit)
hist(posterior$text_profit)
```
```{r}
# Add the column posterior$profit_diff
posterior$profit_diff <- posterior$video_profit - posterior$text_profit

# Visualize posterior$profit_diff
hist(posterior$profit_diff)

# Calculate a "best guess" for the difference in profits
median(posterior$profit_diff)

# Calculate the probability that text ads are better than video ads
sum(posterior$profit_diff < 0) / length(posterior$profit_diff)
head(posterior)
```

Question
So it seems that the evidence does not strongly favor neither text nor video ads.

But if forced to choose at this point, what would you choose?

Text ads


```{r}
# Simulate from a Poisson distribution and visualize the result
x <- rpois(n = 10000, lambda = 3)
hist(x)

# Simulate from a Poisson distribution and visualize the result
x <- rpois(n = 10000, lambda = 11.5)
hist(x)

mean(x >= 15)
```

Question
Is it likely that you will break even on a cloudy day, or should you stay at home?
Let's stay at home and wait for a sunny day.

```{r}
# Change this model so that it uses a Poisson distribution instead
n_draws <- 100000
mean_clicks <- runif(n_draws, min = 0, max = 80)
n_visitors <- rpois(n = n_draws, lambda = mean_clicks)
                     
prior <- data.frame(mean_clicks, n_visitors)
posterior <- prior[prior$n_visitors == 19, ]
hist(prior$mean_clicks)
hist(posterior$mean_clicks)
```

Question
The model is complete! Like before you could now calculate credible probability intervals using quantile or calculate the probability of getting more than, say, 15 clicks next day. But just looking at the posterior probability distribution:

What range could you expect the mean number of daily clicks to be in?

12 to 28 daily clicks on average
