---
title: "Fundamentals of Bayesian Data Analysis in R"
subtitle: "More parameters, more data, and more Bayes"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(purrr)
library(tibble)
library(forcats)
library(ggplot2)
library(ggjoy)
library(lattice)
```

```{r}
temp   <- c(19, 23, 20, 17, 23)
temp_f <- c(66, 73, 68, 63, 73)
like <- dnorm(x = temp, mean = 20, sd = 2)
like
prod(like)
log(like)


```

```{r}
x <- c(3164, 3362, 4435, 3542, 3578, 4529)

# Explore using rnorm and dnorm
mu <- 3700
sigma <- 550

weight_distr <- rnorm(n = 100000, mean = mu, sd = sigma)
hist(weight_distr, 60, xlim = c(0, 6000), col = "lightgreen")

weight <- seq(from=0, to = 6000, by = 100)

likelihood <- dnorm(x = weight, mean = mu, sd = sigma)

plot(weight, likelihood, type = "h")
```

```{r}
n_ads_shown <- 100
n_visitors <- 13

temp <- c(19, 23, 20, 17, 23)
mu <- seq(8, 30, by = 0.5)
sigma <- seq(0.1, 10, by = 0.3)
pars <- expand.grid(proportion_clicks = proportion_clicks)
pars$prior <- dunif(pars$proportion_clicks, min = 0, max = 0.2) 
pars$likelihood <- dbinom(n_visitors, 
    size = n_ads_shown, prob = pars$proportion_clicks)
pars$probability <- pars$likelihood * pars$prior
pars$probability <- pars$probability / sum(pars$probability)

plot(pars, pch=19)

```

```{r}
# The IQ of a bunch of zombies
iq <- c(55, 44, 34, 18, 51, 40, 40, 49, 48, 46)
# Defining the parameter grid
pars <- expand.grid(mu = seq(0, 150, length.out = 100), 
                    sigma = seq(0.1, 50, length.out = 100))
# Defining and calculating the prior density for each parameter combination
pars$mu_prior <- dnorm(pars$mu, mean = 100, sd = 100)
pars$sigma_prior <- dunif(pars$sigma, min = 0.1, max = 50)
pars$prior <- pars$mu_prior * pars$sigma_prior
# Calculating the likelihood for each parameter combination
for(i in 1:nrow(pars)) {
  likelihoods <- dnorm(iq, pars$mu[i], pars$sigma[i])
  pars$likelihood[i] <- prod(likelihoods)
}
# Calculate the probability of each parameter combination
pars$probability <- pars$likelihood * pars$prior / sum(pars$likelihood * pars$prior)

levelplot(probability ~ mu * sigma, data = pars)
```

What is the most probable mean (mu) of the IQ of Zombies? Roughly how uncertain is the estimate of mu?

A mean IQ of 42 ± 10


```{r}
head(pars)
sample_indices <- sample( nrow(pars), size = 10000,
    replace = TRUE, prob = pars$probability)
head(sample_indices)

# Sample from pars to calculate some new measures
pars_sample <- pars[sample_indices, c("mu", "sigma")]

# Visualize pars_sample
hist(pars_sample$mu)

# Calculate the 0.025, 0.5 and 0.975 quantiles of pars_sample$mu
quantile(pars_sample$mu , probs = c(0.025, 0.5, 0.975 ))
```


Question
The 50% quantile you just calculated is the same as the median and a good candidate for a "best guess" for the mean IQ of a zombie, and the 2.5% and 97.5% quantiles form a 95% credible interval.

When submitting this result to The Journal of Zombieology, which of the following sentence should we put into the result section?
Keep in mind that as pars_sample$mu is a random sample, the numbers you've calculated might differ slightly from the possible answers below.


We estimate the mean zombie IQ to be 42 (95% CI: [35, 50])


```{r}
head(pars_sample)
pred_iq <- rnorm(10000, mean = pars_sample$mu, 
                 sd = pars_sample$sigma)
# Visualize pred_iq
hist(pred_iq)

# Calculate the probability of a zombie being "smart" (+60 IQ)
mean(pred_iq >= 60)
```

```{r}
library(BEST)
iq <- c(55, 44, 34, 18, 51, 40, 40, 49, 48, 46)
fit <- BESTmcmc(iq)
fit
plot(fit)
```


Question
Zombies with an IQ of 60 or more are of moderate intelligence, and much more dangerous! (They can open doors!)

How nervous should you be that the next zombie you meet will be one of these smart zombies?

The risk is relatively low but still very real. I always carry my zombie repellent spray!


Question
This plot shows the posterior probability distribution over the difference in means between iq_brains and iq_regular. On top of this you get:

(1) The mean of the posterior as a "best guess" for the difference.

(2) A 95% credible interval (called a 95% Highest Density Interval in the plot).

(3) The amount of probability below and above zero difference.

```{r}
# The IQ of zombies on a regular diet and a brain based diet.
iq_brains <- c(44, 52, 42, 66, 53, 42, 55, 57, 56, 51)
iq_regular <- c(55, 44, 34, 18, 51, 40, 40, 49, 48, 46)

# Calculate the mean difference in IQ between the two groups
mean(iq_brains) - mean(iq_regular)

# Fit the BEST model to the data from both groups
library(BEST)
best_posterior <- BESTmcmc(iq_brains, iq_regular)

# Plot the model result
plot(best_posterior)
```

What would be a reasonable conclusion to draw from this analysis?

There is some evidence that eating brains makes zombies smarter, but it's uncertain by how much.


```{r}
# The IQ of zombies given a regular diet and a brain based diet.
iq_brains <- c(44, 52, 42, 66, 53, 42, 55, 57, 56, 51)
iq_regular <- c(55, 44, 34, 18, 51, 40, 40, 49, 48, 150)

# Modify the data above and calculate the difference in means
 mean(iq_brains) - mean(iq_regular)

# Fit the BEST model to the modified data and plot the result
library(BEST)
best_posterior <- BESTmcmc(iq_brains, iq_regular)
plot(best_posterior)
```


Question
Looking at the plot, we see that the mutant zombie data point has made BEST more uncertain to some degree. But since BEST is robust to outliers, it still estimates that brain-eating zombies are more likely to have a higher IQ than zombies on a regular diet.

What conclusion should we draw?

There is weak evidence that eating brains make zombies smarter. And we should be better at screening for mutant zombies when doing experiments.
