---
title: "Fundamentals of Bayesian Data Analysis in R"
subtitle: "Bayesian inference with Bayes' theorem"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Calculate the probability of drawing any of the four aces
prob_to_draw_ace <- 4/52

# Calculate the probability of picking four aces in a row
prob_to_draw_four_aces <- 4/52 * 3/51 * 2/50 * 1/49
```

Simulating P(n_visitors = 13 | prob_success = 10%)


```{r}
 n_visitors <- rbinom(n = 100000, size = 100, prob = 0.1)
sum(n_visitors == 13) / length(n_visitors)
dbinom(13, size = 100, prob = 0.1)
```

Calculating P(n_visitors = 13 or n_visitors = 14 | prob_success = 10%)

```{r}
 dbinom(13, size = 100, prob = 0.1) + dbinom(14, size = 100, prob = 0.1)
```
 
Calculating P(n_visitors | prop_success = 10%)


```{r}
n_visitors = seq(0, 100, by = 1)
probability <- dbinom(n_visitors, size = 100, prob = 0.1)
plot(n_visitors, probability, type = "h")
```

```{r}
plot(n_visitors, probability, type = "h")

```
The Uniform distribution

```{r}
 x <- runif(n = 100000, min = 0.0, max = 0.2)
hist(x)
```

The Uniform distribution
The d-version of runif is dunif:
 
```{r}
dunif(x = 0.12, min = 0.0, max = 0.2)
```
 
```{r}
 # Rewrite this code so that it uses dbinom instead of rbinom
n_ads_shown <- 100
proportion_clicks <- 0.1
n_visitors <- rbinom(n = 99999, 
    size = n_ads_shown, prob = proportion_clicks)
prob_13_visitors <- sum(n_visitors == 13) / length(n_visitors)
prob_13_visitors

# Rewrite this code so that it uses dbinom instead of rbinom
n_ads_shown <- 100
proportion_clicks <- 0.1
prob_13_visitors <-  dbinom(13, size = n_ads_shown, prob = proportion_clicks)
prob_13_visitors
```
 
```{r}
# Explore using dbinom to calculate probability distributions
n_ads_shown <- 100
proportion_clicks <- 0.1
n_visitors <- seq(0,100,1)
prob <- dbinom(n_visitors, 
    size = n_ads_shown, prob = proportion_clicks)
prob
plot(n_visitors, prob, type = "h")

# Explore using dbinom to calculate probability distributions
n_ads_shown <- 100
proportion_clicks <- seq(0,1,.01)
n_visitors <- 13 # seq(0,100,1)
prob <- dbinom(n_visitors, 
    size = n_ads_shown, prob = proportion_clicks)
prob
plot(proportion_clicks, prob, type = "h")
```
 
 Question
You have now almost done some Bayesian computation. The plot you just produced almost looks like it shows the probability distribution over different values of proportion_clicks, but it does not. For one, the values in prob do not sum up to one. What you have calculated is the likelihood of different values of proportion_clicks to result in n_visitors = 13.

Looking at the plot, what value of proportion_clicks seems to give the maximum likelihood to produce n_visitors = 13?

proportion_clicks = 0.13


```{r}
n_ads_shown <- 100
proportion_clicks <- seq(0, 1, by = 0.01)
n_visitors <- seq(0, 100, by = 1)
pars <- expand.grid(proportion_clicks = proportion_clicks,
                    n_visitors = n_visitors)
pars$prior <- dunif(pars$proportion_clicks, min = 0, max = 0.2)
pars$likelihood <- dbinom(pars$n_visitors, 
    size = n_ads_shown, prob = pars$proportion_clicks)
pars$probability <- pars$likelihood * pars$prior
pars$probability <- pars$probability / sum(pars$probability)
# Condition on the data 
pars <- pars[pars$n_visitors == 6,]
# Normalize again
pars$probability <- pars$probability / sum(pars$probability)
# Plot the posterior pars$probability
plot(pars$proportion_clicks, pars$probability, type = "h")

# Simplify the code below by directly conditioning on the data
n_ads_shown <- 100
proportion_clicks <- seq(0, 1, by = 0.01)
n_visitors <- 6 #seq(0, 100, by = 1)
pars <- expand.grid(proportion_clicks = proportion_clicks,
                    n_visitors = n_visitors)
pars$prior <- dunif(pars$proportion_clicks, min = 0, max = 0.2)
pars$likelihood <- dbinom(pars$n_visitors, 
    size = n_ads_shown, prob = pars$proportion_clicks)
pars$probability <- pars$likelihood * pars$prior
pars$probability <- pars$probability / sum(pars$probability)
#pars <- pars[pars$n_visitors == 6, ]
#pars$probability <- pars$probability / sum(pars$probability)
plot(pars$proportion_clicks, pars$probability, type = "h")

```

```{r}
n_draws <- 100000
mean_clicks <- runif(n_draws, min = 0, max = 80)
n_visitors <- rpois(n = n_draws, mean_clicks)
```

