---
title: "Untitled"
author: "Marjorie Blanco"
date: "11/3/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

```

# PROJECT: EXPLORING THE KAGGLE DATA SCIENCE SURVEY

## Welcome to the world of data science
Throughout the world of data science, there are many languages and tools that can be used to complete a given task. While you are often able to use whichever tool you prefer, it is often important for analysts to work with similar platforms so that they can share their code with one another. Learning what professionals in the data science industry use while at work can help you gain a better understanding of things that you may be asked to do in the future.

In this project, we are going to find out what tools and languages professionals use in their day-to-day work. Our data comes from the Kaggle Data Science Survey which includes responses from over 10,000 people that write code to analyze data in their daily work.

```{r}
# This sets plot images to a nice size.
options(repr.plot.width = 4, repr.plot.height = 4)

# Loading in the ggplot2 package
# .... YOUR CODE FOR TASK 1 ....
library(ggplot2)
```

## Warming up: drawing points on a circle
There are many ways to represent data with ggplot2: from simple scatter plots to more complex ones, such as violin plots. The functions that start with geom_ define how the plot is shown. In this notebook, we will only work with geom_point which plots points in two dimensions. We just need a dataset with two variables, let's call them x and y.

Let's start by drawing 50 points on a circle of radius 1. As every (x, y) point should be in the unit circle, it follows that x² + y² = 1. We can get this using the superfamous Pythagorean trigonometric identity which states that sin²(θ) + cos²(θ) = 1 for any real number θ.

```{r}
t <- seq(0, 2*pi, length.out = 50)
x <- sin(t)
y <- cos(t)
df <- data.frame(t, x, y)

# Make a scatter plot of points in a circle
p <- ggplot(df, aes(x, y))
p + geom_point()
```

## Make it harmonious with the Golden Angle
Plants arrange their leaves in spirals. A spiral is a curve starts from the origin and moves away from this point as it revolves around it. In the plot above all our points are at the same distance from the origin. A simple way to arrange them in a spiral is to multiply x and y by a factor which increases for each point. We could use t as that factor, as it meets these conditions, but we will do something more harmonious. We will use the Golden Angle:

Golden Angle = π(3 − √5)

This number is inspired by the Golden Ratio, one of the most famous numbers in the history of mathematics. Both the Golden Ratio and the Golden Angle appear in unexpected places in nature. Apart of flower petals and plant leaves, you'll find them in seed heads, pine cones, sunflower seeds, shells, spiral galaxies, hurricanes, etc.

It's time to spiralize!

```{r}
# Defining the number of points
points <- 500

# Defining the Golden Angle
angle <- pi * 3 - sqrt(5)
angle <- pi * (3 - sqrt(5))

t <- (1:points) * angle
x <- sin(t)
y <-cos(t)
df <- data.frame(t, x, y)

# Make a scatter plot of points in a spiral
p <- ggplot(df, aes(x*t, y*t))
p + geom_point()
```


## Remove everything unnecessary
Apart from data, a plot includes many other components that define its final appearance. Our previous plot contains:

a background filled with grey color.
a grid of horizontal and vertical white lines in the background.
ticks along the axis.
a title on each axis.
text along axes to label marks.
Art does not get along with most of these elements, so it's time to move to action.

```{r}
df <- data.frame(t, x, y)

# Make a scatter plot of points in a spiral
p <- ggplot(df, aes(x*t, y*t))
p + geom_point() +
  # .... YOUR CODE FOR TASK 4 ....
  theme(panel.grid =  element_blank(), 
        axis.ticks =  element_blank(), 
        axis.title = element_blank(),
        axis.text =  element_blank(),
        plot.background = element_rect(fill="white"))
```


## A bit of makeup: size, color and transparency
Our drawing starts to look like a plant, but we can do it much better. By changing color, transparency (also called alpha), and size of the points, the image will become more appealing.

```{r}
p <- ggplot(df, aes(x*t, y*t))
p + # .... YOUR CODE COPIED FROM TASK 4 ....
  
  
  # .... YOUR CODE FOR TASK 4 ....
  geom_point(size = 8, alpha = 0.5, color = "darkgreen") +
  theme(panel.grid =  element_blank(), 
        axis.ticks =  element_blank(), 
        axis.title = element_blank(),
        axis.text =  element_blank(),
        panel.background = element_rect(fill="white"))
```

## Play with aesthetics: the dandelion
Until now, all points have the same appearance (size, color, shape, and alpha). Sometimes you will want to make the appearance of the points dependent on a variable in your dataset. Now we will make size variable. We will also change the shape of points. Although we won't be able to blow on it, the resulting image should remind you of a dandelion.

```{r}
p <- ggplot(df, aes(x*t, y*t))
p + # .... YOUR CODE COPIED FROM TASK 5 ....
geom_point(alpha = 0.5, color = "black", shape =8, aes(size = t), data = df) +
theme(panel.grid =  element_blank(), 
      axis.ticks =  element_blank(), 
      axis.title = element_blank(),
      axis.text =  element_blank(),
      legend.position  = 'none',
      panel.background = element_rect(fill="white"))
```

## Put all it together: the sunflower
Plants not only use the Golden Angle to arrange leaves. It is also found in the arrangement of sunflower seeds. We don't need anything new to draw a sunflower; we just need to combine some of the things we already know.

```{r}
p <- ggplot(df, aes(x*t, y*t))
p + # .... YOUR CODE COPIED FROM TASK 6 ....
geom_point(alpha = 0.5, color = "yellow", shape =17, aes(size = t), data = df) +
theme(panel.grid =  element_blank(), 
      axis.ticks =  element_blank(), 
      axis.title = element_blank(),
      axis.text =  element_blank(),
      legend.position  = 'none',
      panel.background = element_rect(fill="darkmagenta"))
```


##  What if you modify the angle?
These patterns are very sensitive to the angle between the points that form the spiral; small changes to the angel can generate very different images. Let's look at an example of that


```{r}
angle <- 2
points <- 2000

t <- (1:points)*angle
x <- sin(t)
y <- cos(t)

df <- data.frame(t, x, y)

p <- ggplot(df, aes(x*t, y*t))
p + # .... YOUR CODE COPIED FROM TASK 7 ....
geom_point(alpha = 0.1, color = "yellow", shape = 1, aes(size = t), data = df) +
theme(panel.grid =  element_blank(), 
      axis.ticks =  element_blank(), 
      axis.title = element_blank(),
      axis.text =  element_blank(),
      legend.position  = 'none',
      panel.background = element_rect(fill="white"))
```

## 

```{r}
angle <- 13*pi/180
points <- 2000

t <- (1:points)*angle
x <- sin(t)
y <- cos(t)

df <- data.frame(t, x, y)
# .... COPY ALL THE CODE FROM TASK 8 ....
p <- ggplot(df, aes(x*t, y*t))
p + # .... YOUR CODE COPIED FROM TASK 7 ....
geom_point(alpha = 0.1, color = "magenta4", shape =1, size = 80, aes(size = t), data = df) +
theme(panel.grid =  element_blank(), 
      axis.ticks =  element_blank(), 
      axis.title = element_blank(),
      axis.text =  element_blank(),
      legend.position  = 'none',
      panel.background = element_rect(fill="white"))
```
